\chapter{Introduction}\label{chap:intro}
\section{Context of the Thesis}\label{sec:context}
In the last years, the increasing availability of connected devices has progressively shifted the data generation process from centralized infrastructures towards large populations of edge devices, including smartphones, wearables, vehicles and sensors.
According to~\citeA{cisco2023annual}, the number of networked devices and connections was expected to reach 29.3 billion by 2023, corresponding to more than three times the global population, and approximately 3.6 devices per person. Each of these devices represents a valuable data generation source, collecting information about daily interactions between individuals, devices and digital services.

In this context, Machine Learning (ML) emerges as an incredibly powerful technology to automatically extract patterns from enormous amount of data. In general, access to more data allows to capture richer statistical regularities, improve generalization, and adapt to a wider range of users, environments and tasks~\cite{banko-brill-2001-scaling, shalev2014understanding, hestness2017deeplearningscalingpredictable, kaplan2020scalinglawsneurallanguage}. 

However, having access to more data does not necessarily imply that these data can be freely collected or used. In many domains, data are subject to legal, ethical, technical, and social constraints~\cite{eu_gdpr_2016, hiroshima_process_guiding_principles_2023, eu_ai_act_2024, Buick_copyright, eu_ehds_2025, california_ccpa_cpra}. Medical records are protected because they describe a person’s health status. Financial data may reveal income, consumption patterns, debt, or economic vulnerability. Text, images, and user-generated content may be protected by copyright or by terms of service. Location traces may disclose where a person lives, works, studies, or receives medical care. Even apparently harmless digital traces can become sensitive when aggregated over time or combined with other sources of information \cite{netflix_attack}. Data are therefore not only an input to optimization procedures but are also representations of individuals, their habits, preferences, behaviors, and social relations. 

This observation raises a broader question about control. If data generated by personal devices can reveal intimate aspects of people’s lives, who has access to such data, and under which conditions? In practice, large private companies often occupy a privileged position in the digital ecosystem. They design the devices, operating systems, applications, cloud services, and platforms through which data are generated, stored, and processed. As a consequence, they may acquire detailed knowledge about users’ behavior, preferences, movements, purchases, communications, and interactions. This knowledge can be used in many useful ways to improve people’s daily living, but at the same time, it can also be used for profiling, targeting, ranking, persuasion, facial recognition, or other forms of automated decision-making that are not always transparent to the individuals concerned and are beneficial only for few peoples \cite{pasquale_black_box_society_2015, veliz_privacy_is_power_2021}.

The question is therefore not only whether data are useful, but whether the entities that collect and process them can be trusted. Recent enforcement actions involving major technology companies suggest that privacy risks often arise not only from external attacks or accidental data breaches, but also from ordinary platform practices.

The Cambridge Analytica scandal showed how personal information collected through a digital platform could be repurposed for political profiling and targeted persuasion, raising public concern about consent, transparency, and the secondary use of personal data \cite{ico_democracy_disrupted_2018, ico_data_analytics_political_campaigns_2018}. This episode is part of a broader pattern of documented privacy controversies involving major technology companies \cite{cnil_google_gdpr_2019, ftc_youtube_coppa_2019, ftc_twitter_security_data_ads_2022, ftc_amazon_alexa_coppa_2023, texas_ag_google_privacy_2025}. These cases suggest that privacy risks do not arise only from external breaches, but also from the ordinary ways in which platform operators collect, retain, combine, and repurpose personal data.

Similar concerns arise in machine learning, where the demand for large and diverse training datasets creates strong incentives to collect and reuse data at scale. Controversies involving biometric systems trained on scraped or improperly obtained facial images \citeweb{noyb_clearview_criminal_complaint_2025}, copyrighted material used for model training \citeweb{reisner_ai_pirated_books_problem_2025}, and wearable devices capturing data from both users and bystanders \citeweb{svd_meta_ai_glasses_privacy_2026} illustrate how the search for training data can blur the boundary between legitimate data use and unauthorized appropriation.


These concerns highlight two obstacles to the conventional centralized learning paradigm. The first is practical. Data are increasingly generated across large populations of devices, so collecting them into a single repository may require substantial communication and storage costs, making centralized collection inefficient and difficult to maintain at scale. 

The second obstacle concerns privacy and trust. Even when centralization is technically feasible, transferring raw records to a server increases the exposure of sensitive information and requires users or institutions to trust the entity that collects, stores, and processes the data. The problem is therefore not only how to learn from more data, but how to learn from distributed data without requiring all raw records to be pooled under the control of a single actor.

\section{Federated Learning} \label{sec:fed_learning}

Federated Learning (FL) has emerged as a response to this challenge \cite{konecny2015federatedoptimizationdistributedoptimizationdatacenter, pmlr-v54-mcmahan17a}. Instead of collecting raw data in a central repository, FL moves the training procedure closer to the data sources. Clients keep their datasets locally, compute model updates on their own devices, and communicate only these updates to a central server, whose role is to coordinate the training process and aggregate the received information. In the federated setting, the information exchanged with the server is limited, at least by design, to the updates required for collaborative model training.

This design addresses the two obstacles discussed above. First, it reduces the practical burden of centralized data collection: transmitting model updates can be substantially more efficient than transferring large volumes of raw data to a central storage infrastructure. Second, by avoiding the direct transmission of raw records, FL is naturally aligned with the principle of \textit{data minimization}, according to which only data that are directly relevant and necessary to accomplish a specified purpose should be collected \cite{eu_gdpr_2016, eu_regulation_2018_1725, TRUONG2021102402}.

This second property has made FL particularly attractive in privacy-sensitive domains, including healthcare and finance \cite{Rieke2020, ZHANG2021106775, 10.1145/3298981}. In such settings, FL appears to support the idea of \textit{privacy by design}: privacy considerations are incorporated into the learning architecture itself, rather than being treated only as an external constraint. This intuition is also reflected in the original formulation of FL, where the authors observe that the updates exchanged during training ``will generally contain much less information than the raw training data'' \cite{pmlr-v54-mcmahan17a}.

\subsection{Problem Formulation}

We now formalize the federated learning setting, following the original formulation of \cite{pmlr-v54-mcmahan17a}. Let $\mathcal{U}=\{1,\dots,U\}$ denote a finite set of participants, commonly referred to as clients, users, or devices, coordinated by a central server. The objective is to train a single global machine learning model with parameters $\boldsymbol{\theta}\in\mathbb{R}^d$. Each client $u\in\mathcal{U}$ holds a local dataset $\mathcal{D}_u=\{(\mathbf{x}_i^u, y_i^u)\}_{i=1}^{N_u}$, where $N_u = |\mathcal{D}_u|$ denotes the number of samples available at client $u$. Local datasets may differ both in size and in their underlying data distributions, reflecting the statistical heterogeneity that is typical of federated learning scenarios.

The goal of federated learning is to minimize a global empirical risk obtained by aggregating the local objectives of the clients:

\begin{equation}
\min_{\boldsymbol{\theta} \in \mathbb{R}^d}
\mathcal{L}(\boldsymbol{\theta})
=
\sum_{u=1}^U p_u \mathcal{L}_u(\boldsymbol{\theta}),
\end{equation}
where $\mathcal{L}_u(\boldsymbol{\theta})$ denotes the local objective of client $u$ evaluated on its local dataset $\mathcal{D}_u$, and $p_u$ is the aggregation weight associated with client $u$. A common choice is $p_u = 1/U$ for all $u$, which assigns the same weight to each client. Alternatively, one may set $p_u = \frac{|\mathcal{D}_u|}{|\mathcal{D}|}$, where
$\mathcal{D} = \bigcup_{u=1}^U \mathcal{D}_u$, so that clients are weighted proportionally to the size of their local datasets.

The local objective is typically written as an Empirical Risk Minimization (ERM) problem:
\begin{equation}
\mathcal{L}_u(\boldsymbol{\theta})
=
\frac{1}{N_u}
\sum_{i=1}^{N_u}
\ell(\boldsymbol{\theta};\mathbf{x}_i^u,y_i^u),
\end{equation}
where $\ell(\boldsymbol{\theta};\mathbf{x}_i^u,y_i^u)$ is the loss incurred by the model with parameters $\boldsymbol{\theta}$ on the $i$-th sample of client $u$.

The original algorithm proposed to solve this optimization problem is Federated Averaging (FedAvg) \cite[Algorithm~1]{pmlr-v54-mcmahan17a}, an iterative procedure performed over $T$ communication rounds. Its main steps are summarized in Algorithm~\ref{alg:fedavg}. As discussed above, the training process is coordinated by a central server, which initializes the global model $\model$ (line 1). At each communication round $t$, the server selects a subset of participating clients, denoted by $\mathcal{U}^t \subseteq \mathcal{U}$ (line 3), and sends them the current global model (line 4). After receiving the model, each selected client performs local training on its private dataset (line 6), typically by running stochastic gradient descent for a fixed number of local epochs $E$. Once local training is completed, each client sends back an updated model $\model_u^{t+1}$ (line 7). The server then aggregates the received client models to obtain the next global model $\model^{t+1}$ (line 8).

FedAvg represents the canonical algorithmic formulation of federated learning. Since its introduction, many variants have been proposed, modifying different components of the protocol, including the client optimizer, the server aggregation rule, the client sampling strategy, and the communication pattern \cite{reddi2021adaptivefederatedoptimization}. This line of work shows that federated learning should not be understood as a single fixed algorithm, but rather as a flexible framework for distributed optimization under data locality constraints. At the same time, this flexibility is needed because the federated setting departs from the assumptions of conventional centralized training. Clients may hold data drawn from different distributions \cite{Li2020On, marfoq2021federated, wang_client, scaffold}, operate under different computational \cite{li2019fedmdheterogenousfederatedlearning,FjORD} and network conditions \cite{caldas2019expandingreachfederatedlearning, chen_efficient, ALMANIFI2023100742}, participate intermittently \cite{ribeiro_client_corr, rodio_clients, sun_participation}, and have different privacy requirements \cite{brendan2018learning, Fu_dp_dl_review}. As a result, federated learning introduces a set of challenges that are either absent or less pronounced in centralized learning, including statistical and system heterogeneity \cite{ye_review_het_fl}, communication efficiency \cite{ALMANIFI2023100742}, fairness \cite{Vucinich_fairness, Shi_fairness}, personalization \cite{tan_personalization}, robustness \cite{Uddin_robustness}, and privacy \cite{fu_privacy}.

\begin{algorithm2e}[ht]
\caption[FedAvg: Federated Averaging]{FedAvg: Federated Averaging}
\label{alg:fedavg}
\KwInput{clients $\mathcal{U}$ with local datasets $\{\mathcal{D}_u\}_{u=1}^{U}$; communication rounds $T$; local epochs $E$; mini-batch size $B$; learning rate $\eta$}
\KwOutput{final global model $\boldsymbol{\theta}^T$}
\textbf{server} initializes $\boldsymbol{\theta}^1$\;

\For{$t = 1,\dots,T$}{
    \textbf{server} selects $\mathcal{U}^t \subseteq \mathcal{U}$ clients\;
    \textbf{server} broadcasts $\boldsymbol{\theta}^t$ to the clients in $\mathcal{U}^t$\;
    \ForEach{$u \in \mathcal{U}^t$ \textbf{in parallel}}{
        $\boldsymbol{\theta}^{t+1}_{u}
        \leftarrow
        \operatorname{ClientUpdate}(u,\boldsymbol{\theta}_t,E,B,\eta)$\;
        \textbf{client} sends $\boldsymbol{\theta}^{t+1}_{u}$ to the \textbf{server}\;
    }
    \textbf{server} aggregates $\displaystyle
    \boldsymbol{\theta}^{t+1}
    \leftarrow
    \sum_{u \in \mathcal{U}^t}
    \frac{N_u}{\sum_{v \in \mathcal{U}^t} N_v}
    \boldsymbol{\theta}^{t+1}_{u}$\;
}
\BlankLine
\Procedure{$\operatorname{ClientUpdate}(u,\boldsymbol{\theta},E,B,\eta)$}{
    \textbf{client} splits $\mathcal{D}_u$ into mini-batches $\mathcal{B}_u$ of size $B$\;
    \For{$e = 1,\dots,E$}{
        \ForEach{mini-batch $b \in \mathcal{B}_u$}{
            \textbf{client} updates $\displaystyle
            \boldsymbol{\theta}
            \leftarrow
            \boldsymbol{\theta}
            -
            \eta
            \nabla_{\boldsymbol{\theta}}
            \left(
            \frac{1}{|b|}
            \sum_{(\mathbf{x},y)\in b}
            \ell(\boldsymbol{\theta};\mathbf{x},y)
            \right)$\;
        }
    }
    \Return{$\boldsymbol{\theta}$}\;
}
\end{algorithm2e}


\subsection{Challenges in Federated Learning}
% TODO: riprendi da qui e aggiungi citazioni
\paragraph{Heterogeneity}
Unlike centralized training, where data are typically collected in a single infrastructure and processed under relatively uniform computational conditions, federated learning operates over a population of clients that contribute both data and computation to the training process. This creates two closely related forms of heterogeneity: \textit{statistical heterogeneity}, which concerns differences in the local data distributions, and \textit{system heterogeneity}, which concerns differences in the computational and communication capabilities of the clients.


System heterogeneity arises from differences in clients' hardware, energy availability, memory, bandwidth, and availability over time. These factors determine whether a client can participate in a given communication round, how much local computation it can perform, and how quickly it can return its update to the server. Since such conditions are often outside the direct control of the server, client participation in federated learning is typically partial, intermittent, and uneven across the population.

This uneven participation can affect the training process in several ways. If some clients are more likely to participate than others, the aggregated updates may overrepresent specific regions of the client population, potentially introducing bias in the global model \cite{wang_client}. Moreover, temporal or spatial correlations in client availability can affect convergence and increase the variability of model performance across rounds. System heterogeneity also exacerbates the \textit{straggler effect}, where slower clients delay the completion of a training round or must be dropped from aggregation \cite{kairouz_open_problems}.


Statistical heterogeneity arises because each client generates data through its own local interactions, environment, behavior, and preferences. As a result, the local datasets may differ in size, feature distribution, label distribution, or more generally in the relationship between inputs and outputs \cite{kairouz_open_problems, ye_review_het_fl}. Consequently, the data available at each client may not be representative of the overall population, and the local objectives optimized by different clients may point toward different or even conflicting solutions\cite{scaffold}. Training a single global model under these conditions can therefore lead to a model that performs unevenly across clients, or poorly represents clients whose data differ substantially from the majority \citeA{marfoq2021federated}. 

\paragraph{Personalization}
Closely related to statistical heterogeneity, personalization in federated learning addresses the fact that a single global model may not be equally suitable for all clients \cite{marfoq2021federated}. When clients have different data distributions, the model that performs best on average across the population may perform poorly for specific clients or subgroups. One possible solution would be for each client to train an entirely local model, without participating in the federated process. However, in many cases, individual clients do not have enough data to learn accurate and robust models in isolation. Conversely, participating in standard FL allows clients to benefit from the data indirectly contributed by others, but may produce a global model that fails to capture their local distribution.

Personalized federated learning aims to address this trade-off. The goal is to exploit shared information across clients while still adapting the learned model to each client's local data and task. Several approaches have been proposed to achieve this balance. Some methods learn a common global representation and personalize only part of the model, such as the final layers or client-specific heads \cite{fedper, fedrep}. Others formulate personalization as a regularized optimization problem, where each client learns a local model constrained to remain close to a shared global model \cite{pfedme, ditto}. A further line of work uses clustering or mixture-based approaches, in which clients are grouped according to similarities in their data distributions or learned models, and each group is assigned a more specialized model \cite{marfoq2021federated, marfoq_memorization, fedGMM}
\paragraph{Fairness}
Personalization can improve local utility by adapting models to client-specific objectives and data distributions. Nevertheless, federated training should also ensure that this adaptation does not lead to systematically poor performance for under-represented clients, populations, or protected groups.
Fairness is therefore a central challenge in FL, and different deployments may require different fairness guarantees. Accordingly, several notions of fairness have been considered in the literature \cite{Vucinich_fairness,Shi_fairness, salazar_fairness}. For instance, one may require the final model to satisfy group fairness \cite{salazar_fairness}, similarly to centralized machine learning \cite{dwork_fairness, zemel2013learning_fairness}, by avoiding systematic disparities across protected groups defined by attributes such as race, gender, or age. Beyond group fairness, FL also introduces notions of fairness that are specific to the federated setting. A first one is client-level fairness \cite{li2020fairresourceallocationfederated, erraji2026lossgapparityfairness}, which requires the global model to provide comparable utility across clients, so that no client systematically benefits at the expense of others. A second one concerns participation in the training process. Since only a subset of clients is typically selected at each communication round, frequently selected clients may have a stronger influence on the learned model, whereas rarely selected clients may remain poorly represented \cite{huang_fairness, fairness_shi_client_selection}. Finally, fairness may also be understood in terms of contribution: clients that provide more useful updates, data, or computational resources may need to be compensated or incentivized proportionally to their role in the training process \cite{shi_fairness_incentives, Chen_fairness_incentives}. These different fairness notions raise distinct challenges and can interact with privacy in non-trivial ways \cite{chen_privacy_fairness, fairness_fl_rafi}. In particular, enforcing or auditing group fairness often requires access to sensitive attributes or group-level statistics, which may conflict with the privacy-preserving motivation of FL. As discussed in Chapter~\ref{chap:aia}, sensitive attributes used to define fairness groups may themselves be vulnerable to inference attacks during FL training, and different distribution may impact on the vulnearability of the attribute.

\paragraph{Robustness} Machine learning models may be targeted by adversaries during both training and inference. At training time, \textit{poisoning attacks} aim to corrupt the learned model by injecting malicious samples into the training set or perturbing existing ones \cite{biggio_poisoning, cina_poisoning, jagielski_poisoning}. A relevant subclass is represented by \textit{backdoor attacks}, where the model behaves normally on standard inputs but produces an attacker-chosen prediction whenever a specific trigger is present \cite{cina_poisoning}. At inference time, \textit{evasion attacks} manipulate test inputs so as to induce incorrect predictions, often through small but carefully crafted perturbations \cite{biggio_evasion, carlini_wagner_robustness}.

In federated learning, these threats are amplified by the distributed nature of the training process. Since the server aggregates updates produced by clients whose local data and computation are not directly observable, compromised participants can act as \textit{Byzantine} clients \cite{blanchard_byz}, i.e., clients that arbitrarily deviate from the protocol and submit malicious or corrupted updates. Such updates may aim to degrade the overall performance of the global model  \cite{pmlr-v97-bhagoji19a, alie_attack, foe_attack, fang_poisoning}, or implant a backdoor while preserving normal accuracy on clean validation data \cite{pmlr-v108-bagdasaryan20a}. Compared to centralized poisoning, the attacker does not need to contaminate a centrally collected dataset; it can directly influence the optimization process through the updates sent to the server. This has motivated a large body of work on Byzantine-resilient aggregation \cite{blanchard_byz, mhamdi2018hiddenvulnerabilitydistributedlearning} and robust distributed optimization \cite{GuerraouiGP24, Allouah_bounds, Allouah_clipping}. 

\section{Privacy in Federated Learning}
In Section~\ref{sec:fed_learning}, we introduced the main motivations behind federated learning, including the intuitive privacy benefit of keeping data local. However, data locality should not be confused with a privacy guarantee. It reduces the need to transfer raw records to a central repository, but it does not by itself determine what information may still be inferred during training.Privacy depends not only on where data are stored, but also on who can access the information derived from them, how this information is processed, and for which purposes it is used. This is particularly relevant in light of legal frameworks that recognize privacy and data protection as fundamental rights \cite{eu_gdpr_2016,california_ccpa_cpra}. In this context, analyzing privacy risks means asking what information about clients or their data may be exposed to the actors involved in the learning process.
This issue is not specific to federated learning. In centralized machine learning, previous work has shown that trained models may leak information about training data\cite{shokri_mia, aia_fredrikson, carlini_extract_data, feng2024privacy}. In federated learning, however, the training protocol introduces an additional source of leakage as the messages exchanged between clients and server can be intercepted. These updates are computed from local datasets and therefore may carry traces of the private information they are meant to protect. Depending on the adversary's capabilities, such information may be observed passively or amplified through active interference with the protocol.
A large body of work has shown that, without additional safeguards, the messages exchanged during training can reveal private or sensitive information about clients and their local datasets \cite{lyu2020threatsfederatedlearningsurvey, zhao_fl_attacks_review}. The remainder of this section reviews these risks by discussing the main classes of privacy attacks considered in the FL literature, together with the defense strategies proposed to mitigate them.

\subsection{Threat Models}

Before reviewing the main classes of privacy attacks in federated learning, it is useful to clarify the threat models under which such attacks are studied. A threat model specifies the adversary's position in the protocol, the information available to the adversary, the actions the adversary is allowed to perform, and the private information that the adversary aims to infer. In FL, a central distinction is between \textit{honest-but-curious}, or passive, adversaries and \textit{malicious}, or active, adversaries~\cite{Nasr2018ComprehensivePA}.

In the honest-but-curious setting~\cite{honest_but_curious}, the adversary follows the prescribed training protocol but attempts to infer private information from the information naturally revealed during training. This adversary may be, for instance, a curious server that inspects the updates received from clients, or a man-in-the-middle that silently eavesdrops on the communication between clients and the server. Since passive adversaries do not alter the training process, they are generally harder to detect: their attack surface is limited to the messages, models, or updates that are already exchanged as part of the protocol. Nevertheless, this setting can still be highly informative. Given the nature of FL systems, the adversary is often assumed to have white-box knowledge of the system, including the model architecture, current model parameters, loss function, and training algorithm. The passive nature of the adversary therefore refers to its behavior, not necessarily to a lack of knowledge.

By contrast, a malicious adversary deviates from the prescribed protocol in order to amplify privacy leakage. Such an adversary may be a compromised client, a coalition of clients, or, more commonly in the privacy-attack literature \cite{Nasr2018ComprehensivePA, zhao_fl_attacks_review, carletti2025sok}, the federated learning server itself. A malicious server is particularly powerful because it orchestrates the training process, selecting clients and distributing model parameters. This position allows the server to modify the model or the messages sent to targeted clients, with the goal of inducing updates that reveal more information about their local datasets. Active attacks are therefore often stronger than passive attacks, but they also rely on stronger assumptions and may be more detectable if clients are able to verify the received model or monitor deviations from the expected protocol~\cite{mia_survey_bai, carletti2026detectability}.

The black-box, grey-box, and white-box distinction can be used as a complementary axis of the threat model. Black-box access is more typical of post-training attacks, where the adversary only queries a trained model and observes its outputs. Grey-box settings capture intermediate cases, for example when the adversary knows the model architecture but observes only aggregated updates, or when some local training details are unknown. White-box access, instead, is common in training-time FL attacks, where the adversary observes model parameters and updates exchanged during the protocol \cite{Nasr2018ComprehensivePA, zhao_fl_attacks_review}. In this thesis, Chapter~\ref{chap:aia} considers both passive and active adversarial models, whereas the following chapters mainly focus on the privacy risks posed by active adversaries, with particular emphasis on malicious-server scenarios.

\subsection{Membership Inference Attacks}

Membership inference attacks (MIAs) aim to determine whether a given data record is part of the training dataset of a model \cite{shokri2017membershipinferenceattacksmachine,yeom2018privacyriskmachinelearning,mia_survey_bai}. Formally, let $\dataset_{\mathrm{tr}}$ denote the training dataset and let $(\x,y)$ be a candidate record. Given an observation $\mathcal{O}$ available to the adversary, such as the final model, its outputs, gradients, model updates, or a sequence of intermediate models, an MIA is a binary decision rule
\[
    \mathcal{A}_{\mathrm{MIA}}((\x,y),\mathcal{O}) \in \{0,1\},
\]
where $\mathcal{A}_{\mathrm{MIA}}((\x,y),\mathcal{O})=1$ indicates that the adversary infers $(\x,y) \in \dataset_{\mathrm{tr}}$, while $\mathcal{A}_{\mathrm{MIA}}((\x,y),\mathcal{O})=0$ indicates that $(\x,y) \notin \dataset_{\mathrm{tr}}$. MIAs are commonly used as an empirical tool for assessing privacy leakage, since a successful attack shows that the trained model, or the information exchanged during training, contains a distinguishable trace of individual training examples. This leakage can be sensitive even when the adversary learns only membership status. For instance, if a record is found to belong to the training set of a model trained on a specific medical condition, membership may already reveal information about the corresponding individual. Beyond privacy-risk assessment, MIAs have also been used to audit compliance with data deletion or unlearning requirements, and can serve as a preliminary step for stronger attacks, such as model extraction or attacks aiming to infer additional private information \cite{xiao_mia,hu2021sourceinferenceattacksfederated}.

\begin{figure}[tbp]
    \centering
    \resizebox{0.97\textwidth}{!}{%
    \begin{tikzpicture}[
        >=stealth,
        font=\footnotesize,
        stage/.style={
            draw=bleu,
            dashed,
            line width=1pt,
            rounded corners=3pt,
            fill=bleu!2
        },
        card/.style={
            draw=black!70,
            thick,
            rounded corners=3pt,
            fill=white,
            align=center,
            inner sep=3pt
        },
        attackcard/.style={
            draw=vertc!85!black,
            thick,
            rounded corners=3pt,
            fill=vertc!6,
            align=center,
            inner sep=4pt
        },
        arrow/.style={->, line width=1pt, draw=black!75},
        softarrow/.style={->, line width=0.8pt, draw=black!55},
        title/.style={font=\bfseries\small, align=center},
        tinylabel/.style={font=\scriptsize, align=center}
    ]
        \draw[stage] (0,0) rectangle (3.55,3.95);
        \node[title] at (1.78,3.62) {Federated training};

        \draw[stage] (4.35,0) rectangle (7.95,3.95);
        \node[title] at (6.15,3.62) {Adversary's view};

        \draw[stage] (8.65,0) rectangle (13.05,3.95);
        \node[title] at (10.85,3.62) {Inference goal};

        % Server icon.
        \coordinate (server) at (1.78,2.40);
        \draw[bleu, line width=1pt] (1.78,1.93) -- (1.36,2.82) -- (2.20,2.82) -- cycle;
        \draw[bleu, line width=1pt] (1.78,1.93) -- (1.78,3.18);
        \draw[bleu, line width=1pt] (1.50,2.38) -- (2.06,2.38);
        \draw[bleu, line width=1pt] (1.42,2.62) -- (2.14,2.62);
        \draw[bleu, line width=1pt] (1.60,2.86) arc[start angle=210,end angle=-30,radius=0.22];
        \draw[bleu, line width=1pt] (1.44,2.86) arc[start angle=210,end angle=-30,radius=0.38];

        % Client icons.
        \foreach \xphone/\name in {0.54/$u_1$,1.54/$u_2$,2.54/$u_K$} {
            \coordinate (client\xphone) at (\xphone+0.24,1.23);
            \draw[black, line width=0.9pt, rounded corners=2pt, fill=white] (\xphone,0.72) rectangle ++(0.48,0.80);
            \fill[black] (\xphone+0.05,0.82) rectangle ++(0.38,0.60);
            \fill[white] (\xphone+0.24,0.90) circle (0.032);
            \draw[softarrow] (server) -- (\xphone+0.24,1.56);
            \draw[softarrow] (\xphone+0.24,1.56) -- (server);
        }

        \node[card, text width=2.62cm, minimum height=0.42cm] at (1.78,0.21)
            {$\dataset_{\mathrm{tr}}=\bigcup_{k=1}^{K}\dataset_k$};

        % Attacker icon.
        \draw[pink!85!black, line width=0.95pt] (5.78,3.02) -- (6.15,3.40) -- (6.52,3.02);
        \draw[pink!85!black, line width=0.95pt, rounded corners=2pt] (5.87,2.70) rectangle (6.43,3.02);
        \draw[pink!85!black, line width=0.95pt] (5.97,2.70) -- (5.78,2.52) -- (6.52,2.52) -- (6.33,2.70);
        \draw[pink!85!black, line width=0.95pt] (6.15,3.23) circle (0.09);

        \node[card, text width=2.64cm, minimum height=0.78cm] (obs) at (6.15,1.70)
            {$\mathcal{O}$\\[-0.5mm]\scriptsize outputs, losses,\\[-0.4mm]gradients, updates};
        \node[card, text width=2.64cm, minimum height=0.58cm] (cand) at (6.15,0.74)
            {\scriptsize candidate record\\[-0.4mm]$(\x,y)$};

        \draw[arrow] (3.55,2.50) -- node[pos=0.48, above, font=\scriptsize, fill=white, inner sep=1pt] {observed signal} (4.50,2.50) |- (obs.west);
        \draw[arrow] (6.15,2.52) -- (obs.north);

        % Attack rule and goal.
        \node[attackcard, text width=3.24cm, minimum height=0.86cm] (attack) at (10.85,2.70)
            {\scriptsize attack rule\\[0.4mm]$\mathcal{A}_{\mathrm{MIA}}\bigl((\x,y),\mathcal{O}\bigr)$};
        \node[card, text width=1.28cm, minimum height=0.58cm, fill=violet!8] (member) at (10.00,1.42)
            {$1$\\[-0.5mm]\scriptsize member};
        \node[card, text width=1.44cm, minimum height=0.58cm, fill=violet!8] (nonmember) at (11.70,1.42)
            {$0$\\[-0.5mm]\scriptsize non-member};
        \node[card, text width=3.72cm, minimum height=0.58cm, fill=pink!7] (goal) at (10.85,0.42)
            {\scriptsize goal\\[-0.4mm]decide if $(\x,y)\in\dataset_{\mathrm{tr}}$};

        \coordinate (attackObsIn) at (9.23,2.84);
        \coordinate (attackCandIn) at (9.23,2.58);
        \draw[arrow] (obs.east) -- ++(0.30,0) |- node[pos=0.72, above, font=\scriptsize, fill=white, inner sep=1pt] {features} (attackObsIn);
        \draw[arrow] (cand.east) -- ++(0.52,0) |- (attackCandIn);
        \draw[arrow] (attack.south) -- ++(0,-0.32) -| (member.north);
        \draw[arrow] (attack.south) -- ++(0,-0.32) -| (nonmember.north);
    \end{tikzpicture}%
    }
    \caption[Membership inference attack workflow]{Generic membership inference attack workflow. The adversary combines observations from the model or training protocol with a candidate record and tries to decide whether the record belongs to the training data.}
    \label{fig:mia_workflow}
\end{figure}

MIAs were first studied in centralized learning \cite{shokri2017membershipinferenceattacksmachine,yeom2018privacyriskmachinelearning,carlini2022membershipinferenceattacksprinciples}, where the adversary typically interacts with a trained model after deployment and exploits differences between the model's behavior on training and non-training records, for example through prediction confidence, loss, or correctness. In federated learning, however, the attack surface is broader, since the attacker has access not only to the final model, but also to all the intermediate update history. The adversary may be the central server, a participating client, or an external eavesdropper, and may act either passively, by only observing the protocol, or actively, by modifying model parameters or data to amplify membership leakage \cite{mia_survey_bai}.

The first FL-specific MIA was introduced by \citeA{Nasr2018ComprehensivePA}, who proposed a white-box attack in both passive and active settings. In this threat model, the adversary is assumed to know the target model architecture, its parameters, and the hyperparameters needed to run inference. Therefore, for any candidate record $(\x,y)$, the adversary can compute not only the model prediction $f(\x;W)$, but also hidden-layer activations, the loss $L(f(\x;W),y)$, and the gradients of this loss with respect to the parameters of each layer. In the supervised version of the attack, the adversary also has an auxiliary dataset $\dataset_{\mathrm{aux}}$, containing examples with known membership status: some records are known members of the target training set, while others are non-members sampled from the same distribution. These examples provide the labels needed to train the attack model. The attack model then takes, for each candidate record, a collection of white-box features as input, including layer-wise gradients, hidden-layer outputs, the final prediction vector, the one-hot encoding of the true label, and the loss value, collected over multiple training rounds. 
% The attack can be launched by a passive server, which observes individual client updates, or by a passive participant, which observes only aggregated global updates.

Closely related work by \citeA{melis2018exploitingunintendedfeatureleakage} showed that updates exchanged during collaborative learning can leak unintended information about other participants' data, including the presence of specific data points and properties of the local training distribution. Subsequent work on MIAs in FL has reduced the amount of auxiliary information or computation required by the attacker, and has explored different sources of membership signal. For example, \citeA{zhang_gan_mia} use generative adversarial networks to augment the attack data used for shadow training, while \citeA{zari2021efficientpassivemembershipinference} propose a more efficient passive attack based on the evolution of model outputs. This trend-based line of work was extended by attacks that exploit prediction-confidence series \cite{gu_mia}, trajectories of bias parameters \cite{Zhang_mia}, and the temporal evolution of adversarial robustness \cite{liu_mia}. Other approaches remove the need to train a shadow model: \citeA{li2023effective} infer membership from the geometry of gradient updates across consecutive rounds, whereas \citeA{Zhu_mia} formulate the attack as a hypothesis test that leverages updates from multiple clients and communication rounds. 
% MIAs have also been adapted to application-specific FL settings; for instance, \citeA{yuan2023interaction} study interaction-level membership inference in federated recommender systems, where the goal is to determine whether a user interacted with a particular item.

Finally, active attacks show that membership leakage can be amplified when the adversary is allowed to interfere with the protocol. For instance, \citeA{Pichler_mia} modify the parameters of a ReLU-based fully connected layer so that the resulting gradients reveal whether a queried sample was used during training. \citeA{mia_nguyen} show that active MIAs can also be mounted under local differential privacy by crafting malicious model parameters that increase the membership signal observed by the server. Similarly, \citeA{He_mia} use data poisoning to make the model memorize information about specific target classes before launching the inference attack.

In FL, the basic membership question can also be refined to account for the client structure of the training set. Instead of only deciding whether $(\x,y)$ belongs to the global training dataset, an adversary may attempt to infer which client contributed the record. If $\dataset_{\mathrm{tr}}=\bigcup_{k=1}^K \dataset_k$, this source-level variant can be written as
\[
    \mathcal{A}_{\mathrm{source}}((\x,y),\mathcal{O}) \in \{0,1,\ldots,K\},
\]
where $0$ denotes non-membership and $k \in \{1,\ldots,K\}$ indicates that $(\x,y)$ is inferred to belong to the local dataset $\dataset_k$ of client $k$. This refinement is particularly relevant in cross-silo FL, for example when hospitals collaborate to train a model, because identifying the institution that contributed a record may reveal additional sensitive information about the corresponding individual \cite{hu2021sourceinferenceattacksfederated}.
%  A related formulation is subject-level membership inference, where the adversary aims to determine whether the data of a given individual participated in the federation, possibly through multiple records distributed across several institutions \cite{suri2022subject}.
\subsection{Attribute Inference Attacks}\label{sec:aia}

While membership inference attacks reveal whether a record was used during training, attribute inference attacks (AIAs) ask a complementary question: what sensitive information about a partially known record can be inferred from the model or from the training process? AIAs aim to infer a sensitive attribute of a target record from partial knowledge about that record and from information leaked by the learning procedure. Let a data point be decomposed as $(\x^p, s, y)$, where $\x^p$ denotes the public attributes, $s$ is a sensitive attribute, and $y$ is the associated label or target value. Given the public part of the record, possibly the target value, and an observation $\mathcal{O}$ available to the adversary, an AIA can be written as
\[
    \mathcal{A}_{\mathrm{AIA}}(\mathcal{O}, \x^p, y) \mapsto \hat{x}_s .
\]
The observation $\mathcal{O}$ depends on the setting and may correspond, for example, to the output of the trained model, its gradients, model updates, or a sequence of intermediate models.

In the centralized setting, an early and influential instance of this type of attack was introduced by~\citeA{aia_fredrikson} in the medical domain. Their work studies a \textit{model inversion} attack in which the adversary uses access to a trained model, together with auxiliary information about an individual, to infer a missing sensitive attribute. A common model-based formulation of this attack consists in testing different candidate values for the sensitive attribute and selecting the one that makes the completed record most consistent with the trained model. This can be expressed as
\begin{equation}\label{eq:aia_model_based}
    \hat{s}
    \in
    \argmin_{s}
    \ell\bigl(\theta,(\x^p, s, y)\bigr),
\end{equation}
where $\theta$ denotes the trained model and $\ell$ is the loss function used to measure the consistency between the completed record and the model prediction.


The intuition behind this attack is that the learned model may encode information about how the sensitive attribute affects the prediction on records similar to those seen during training. As a result, when the adversary completes a partially known record with the correct sensitive value, the resulting input may be more consistent with the model behavior than completions based on incorrect candidate values. This form of leakage is closely related to the influence of individual features on the learned predictor, as well as to memorization, overfitting, and overlearning phenomena~\cite{yeom2018privacyriskmachinelearning,Song2020Overlearning}. This interpretation, however, requires a careful baseline. Since sensitive attributes may be statistically correlated with the public attributes, an adversary could sometimes infer them even without access to the trained model. For this reason, \citeA{Jayaraman_aia} compare attribute inference attacks with imputation baselines, showing that AIA performance should be interpreted in terms of the additional information provided by model access, rather than only through the raw reconstruction accuracy.

Subsequent work refined both the attack objectives and the adversarial assumptions. For classification models, \citeA{mehnaz2022sensitiveattributesprivatenovel} propose black-box model inversion AIAs that use confidence scores to select the most likely sensitive attribute value, and also introduce a label-only variant that requires only the predicted class. Their work further considers a relaxed setting in which some non-sensitive attributes of the target record are also unknown to the adversary.

In federated learning, the same inference goal can be pursued using information that is not available in the standard centralized setting. Since the server, or an eavesdropping adversary, may observe the model updates exchanged during training, the sensitive attributes can be treated as unknown variables that explain those updates. In particular, given the public part of the client's records and the corresponding labels, the adversary formulates the unknown sensitive attributes as optimization variables and searches for the assignment that makes the gradients computed on the completed records best align with the update observed during federated training~\cite{lyu2021novelattributereconstructionattack,chen2024practical}. More formally, for a target client $u$ and a set of observed rounds $\mathcal{T}$, the adversary can solve

\begin{equation}\label{eq:aia_gradient_based}
  \argmax_{\{s_u(i)\}_{i=1}^{B}} \sum_{t\in \mathcal{T}}\textbf{CosSim} \left(\frac{\partial\ell(\model^t, 
    \{(\x^p(i), s_u(i), y_u^p(i))\})}{\partial \model^t}, \model^t-\model_u^t\right), 
\end{equation}

These attacks have been studied under both honest-but-curious and active adversaries, where the server isolates a target client in the training process to amplify leakage. A different active approach is proposed by~\citeA{Gomes_aia}, where the malicious server modifies selected neurons so that sensitive information about target records is encoded in the sign or structure of the resulting gradient update. 
Existing AIAs in federated learning have mainly been developed for classification tasks, leaving regression comparatively underexplored despite its importance in many practical applications. Moreover, while model-based AIAs were originally studied in the centralized setting, their adaptation to federated learning has received limited attention. Chapter~\ref{chap:aia} addresses this gap by studying model-based attribute inference in federated regression tasks. In particular, it shows how an adversary can approximate the target client's local model and use it to infer sensitive attributes, overcoming limitations of gradient-based attacks in regression in both honest-but-curious and malicious settings. 



\subsection{Data Reconstruction Attacks}\label{sec:dra}
So far, the discussed attacks assumed that the attacker had access to the sample from which he wanted to retrieve some information (MIA) or part of it (AIA). Instead, in Gradient Inversion Attacks (GIAs)---referred to also as Data Reconstruction Attacks (DRAs)--- the goal of the attacker is to reconstruct the entire training sample, by observing the exchanged updates \cite{carletti2025sok,Du_gia_survey}. Existing attacks can be categorized in three main families, namely \textit{optimization-based} attacks, \textit{generative-based}, and \textit{analytical} attacks \cite{carletti2025sok}. Optimization-based attacks can be further refined based on if the goal of the attack is to directly optimize in the input space or in a latent space. 

The attacks discussed so far assume that the adversary already has access either to the record whose membership is to be inferred (MIAs), or to a partially observed record (AIAs). Instead, data reconstruction attacks (DRAs) consider a stronger form of privacy leakage. In this setting, the adversary aims to recover the training data themselves from information exposed by the learning procedure. In federated learning, this information typically consists of the gradients or model updates exchanged between clients and the server. For this reason, these attacks are often referred to as \emph{gradient inversion attacks} (GIAs), although the broader term \emph{data reconstruction attacks} (DRAs) is also used in the literature \cite{carletti2025sok,Du_gia_survey}.

Existing GIAs can be broadly organized into three main families: \emph{optimization-based}, \emph{generative-based}, and \emph{analytical} attacks \cite{carletti2025sok}. Optimization-based attacks can be further distinguished according to whether the reconstruction is performed directly in the input space or in a latent representation. The following paragraphs review these families in more detail.

\paragraph{Optimization-based attacks} 
Optimization-based attacks reconstruct private data by solving an inverse problem. The adversary initializes a candidate reconstructed batch and optimizes its inputs, and possibly its labels, so that the gradients computed on this batch match the update observed during federated training. More formally, for a target client $u$ and a set of observed rounds $\mathcal{T}$, the reconstruction problem can be written as
    \begin{equation}\label{eq:gia_general}
        \{(\hat{\x}_i,\hat{y}_i)\}_{i=1}^{B}
        \in
        \argmin_{\{(\x_i,y_i)\}_{i=1}^{B}}
        \sum_{t\in\mathcal{T}}
        \Dist\left(
            \frac{\partial \ell\left(
                \model^t,
                \{(\x_i,y_i)\}_{i=1}^{B}
            \right)}{\partial \model^t},
            \model^t-\model_u^t
        \right),
    \end{equation}
    where $\Dist(\cdot,\cdot)$ denotes a distance or dissimilarity measure between the gradient computed on the reconstructed batch and the update observed by the adversary. 

The original attack formulation by~\citeA{dlg} instantiated this objective with a Euclidean matching loss. Subsequent work showed that the success of optimization-based GIAs depends not only on solving the inverse problem, but also on how the gradient-matching objective is defined. In particular, \citeA{geiping_ig} showed that image reconstruction can be substantially improved by matching the direction of the observed and reconstructed gradients through a cosine-based objective, and by adding total variation regularization~\cite{rudin_tv} to promote spatially coherent reconstructions and stabilize the optimization.

These early attacks, however, were mainly effective in relatively idealized settings. They often assumed that the target labels were known and achieved high-quality reconstructions only for very small batches. When a client computes gradients over a local batch, the update observed by the server contains information from all samples in that batch, rather than separate per-sample gradients. As the batch size grows, the signal associated with each individual record becomes increasingly difficult to disentangle. A first line of follow-up work therefore focused on making the attack setting more realistic by recovering unknown labels together with the inputs~\cite{zhaoIDLGImprovedDeep2020, maINSTANCEWISEBATCHLABEL2023, gengImprovedGradientInversion2024, yeHighFidelityGradientInversion2024, liTemporalGradientInversion2025a}. In parallel, other works improved reconstruction quality by enriching the optimization objective, either through additional regularization terms~\cite{yinSeeGradientsImage2021, hatamizadehGradViTGradientInversion2022,liE2EGIEndtoEndGradient2023, gengImprovedGradientInversion2024, Fan_fedleak} or through losses designed to better correlate with reconstruction quality~\cite{wangVarianceGradientAlso2022, hongForeseeingReconstructionQuality2024, yeHighFidelityGradientInversion2024, Fan_fedleak}.

Improving the objective alone is often insufficient in large-batch regimes, where the main difficulty is the separation of multiple sample contributions from a single observed update. Some works therefore modify the reconstruction algorithm itself to better handle this regime. For instance, \citeA{kariyappaCocktailPartyAttack2023} formulate gradient inversion as a blind source separation problem, aiming to separate the per-sample gradient contributions contained in the batch gradient. More recently, \citeA{liTemporalGradientInversion2025a} exploit information collected across different training steps and improve reconstruction by combining multiple candidate reconstructions of the same batch, drawing inspiration from robust aggregation techniques.

Another important limitation arises in realistic federated protocols such as FedAvg, where clients perform several local optimization steps before communicating with the server. In this setting, the exhanged update is no longer a single gradient, but the result of an entire local training trajectory. To address this mismatch, \citeA{dimitrovDataLeakageFederated2022} propose to simulate the FedAvg update during reconstruction, so as to better reproduce the client-side computation and recover images from model updates. Later works approximate this local trajectory through surrogate models, using either linear~\cite{Zhu_sme_gia} or non-linear~\cite{xiaNonLinearTrajectoryModeling2025} trajectory models to improve reconstruction from multi-step updates.

Although most optimization-based GIAs were initially designed for images, the same general principle has also been adapted to other data modalities. In the text domain, \citeA{deng-etal-2021-tag-gradient} design TAG, an attack specifically targeting Transformer-based language models. Later, \citeA{lamp_balunovic} propose to guide sentence reconstruction with a language-model prior. These settings still benefit from the fact that reconstruction quality can often be assessed through visual or semantic inspection. In the tabular domain instead, this is not possible: records may contain continuous, categorical, or mixed features, and reconstruction quality may be difficult to assess without access to the ground truth. In this context, \citeA{veroTabLeakTabularData2023} propose the only optimization-based GIA tailored to tabular data, together with an entropy-based heuristic for estimating reconstruction quality.

All the attacks reviewed so far assume an honest-but-curious adversary, who passively observes the updates exchanged during the normal execution of the FL protocol. A stronger threat model considers a malicious adversary, typically the server, who can actively modify the model parameters, the architecture, or parts of the training protocol to induce more informative client updates. This changes the role of optimization-based reconstruction: instead of only improving the inversion procedure, the attacker can first manipulate the training signal so that the observed update reveals more information about selected client samples. For classification tasks, this can be achieved by modifying the classification layer to isolate or amplify the gradients associated with a target class~\cite{GradFiltClasswiseTargeted, wenFishingUserData2022a}. The same principle can also be exploited across multiple communication rounds to progressively focus the leakage on specific samples~\cite{wenFishingUserData2022a}. Alternatively, \citeA{shan2025geminio} use a vision-language model to assist the adversary in crafting a malicious model that increases the loss on target samples described through natural language, thereby making their contribution to the observed update more prominent.

\paragraph{Generative-based attacks} The use of generative models in federated-learning privacy attacks has evolved through several distinct paradigms. Early works employed GANs to approximate class-level distributions in victims’ private data. In the first generateive-based attack proposed by \citeA{hitajDeepModelsGAN2017a}, a malicious participant exploits and actively influences the global model to train a GAN that generates prototypical samples of a target class available only to a victim participant. Instead, in the attack proposed by \citeA{wangInferringClassRepresentatives2018} the server becomes the attacker, and uses the client local updates as proxy for the discriminator. Subsequent studies shifted toward instance-level gradient inversion by constraining the reconstruction search with generative models. \citeA{jeon2021gradient} proposed GIAS, which replaces the direct optimization of the candidate input in Equation~\eqref{eq:gia_general} with an optimization over the output space of a pretrained generator. GIAS first optimizes the latent code and then fine-tunes the generator parameters, thereby compensating for inaccuracies in the pretrained generative prior. For settings in which a suitable pretrained generator is unavailable, the authors further introduced GIML, a meta-learning approach that learns a generative prior across multiple gradient-inversion tasks constructed from updates observed during federated training. GGL~\cite{liAuditingPrivacyDefenses2022a} reconstructs private samples from defense-transformed gradients by searching the latent manifold of a pretrained GAN with adaptive gradient-free optimization. GIFD~\cite{fangGIFDGenerativeGradient2023a} extends this strategy beyond latent-space search by successively optimizing intermediate generator features. GI-SMN~\cite{qian_gismn} subsequently optimizes the latent code of a pretrained StyleGAN-XL while regularizing the generated images through total-variation, $L_2$, and group regulariztions.
AA complementary family of attacks replaces the fixed pretrained prior with a generator optimized during the inversion process~\cite{renGRNNGenerativeRegression2022a,xuCGIRConditionalGenerative2023,zhangGenerativeGradientInversion2023a}. Rather than optimizing the reconstructed images directly, these methods update the generator parameters so that its outputs induce gradients matching the observed client update.

A more recent line of work shifts the reconstruction cost to an offline training phase by learning a reusable inversion model or generative decoder from auxiliary data. Once trained, these models can be applied to newly intercepted updates without solving a separate gradient-matching problem for each reconstruction. LTI~\cite{wu2023learning} follows the most direct approach, training an inversion network on auxiliary sample–gradient pairs to map gradients to their corresponding inputs. Defensive transformations, such as noise addition, pruning, or compression, are incorporated when generating the training gradients, allowing the inversion model to circumvent defenses. With the same goal~\citeA{Yue_gia2023} design an encoder-decoder inversion pipeline that solve the gradient optimization problem in a latent space, and then decodes the reconstruced image using a pretrained decoder.
 FGLA~\cite{xueFastGenerationBasedGradient2023c} instead adopts a hybrid analytical–generative pipeline. It first separates the sample-level representations corresponding to the input of the classifier from the gradients of its fully connected layer, and then feeds these representations to a trained generator that reconstructs the original inputs. Under a malicious-server threat model, Scale-MIA~\cite{shiScaleMIAScalableModel2025} trains a generative decoder on auxiliary data and crafts linear layers in the latent space of the distributed model. The attacker can then recover latent representations from aggregated updates through a closed-form procedure and map them back to the input space using the trained decoder. SEER~\cite{garovHidingPlainSight2024} instead jointly trains the shared model and a secret server-side decoder so that the contribution of individual samples can be separated from the aggregated client gradients in a hidden representation space and then decoded into the corresponding inputs

Finally, diffusion-based attacks use leaked gradients to guide the reverse generation process, as in GGDM~\cite{guFederatedLearningVulnerabilities2024} and gradient-guided conditional diffusion models~\cite{huangGradientGuidedConditionalDiffusion2024, mengEnhancedPrivacyLeakage2025}, while GUIDE~\cite{carlettiGUIDEEnhancingGradient2025} employs diffusion primarily as a post-reconstruction denoiser.A different use of diffusion is proposed by Mjölnir~\cite{liu2025mjolnirbreakingshieldperturbationprotected}, which trains a diffusion model to recover an approximation of the original gradients from perturbed observations before applying a conventional optimization-based reconstruction attack.

Despite these advances, generative-based attacks still face difficulties when reconstructing large batches unless they are combined with complementary techniques, such as analytical gradient separation or active model manipulation. Moreover, their effectiveness has so far been demonstrated predominantly on image data, while their applicability to other modalities remains comparatively limited.

\paragraph{Analytical attacks}
The analytical study of gradient inversion can be traced back
to~\citeA{phong_gia}, who first observed that the input to an affine
layer can be recovered directly from its weight and bias gradients when
these gradients are computed on a single training sample. This observation later became the basis of analytical leakage attacks against fully connected layers followed by ReLU activations. Indeed, ReLU naturally induces sparse gradients, since samples yielding
non-positive pre-activations contribute zero gradient to the corresponding neuron. 

To illustrate this mechanism, consider a fully connected layer with
$N$ neurons, parameterized by a weight matrix
$\W^{a} \in \mathbb{R}^{N \times d}$ and a bias vector
$\b^{a} \in \mathbb{R}^{N}$. For an input
$\x_j \in \mathbb{R}^{d}$, the post-activation output of neuron $i$ is
\begin{equation}
    z^{a}_{i,j}
    =
    \operatorname{ReLU}
    \left(
        \W^{a}_{i}\x_j+b^{a}_{i}
    \right).
\end{equation}
Let $\loss_j$ denote the loss associated with sample
$(\x_j,y_j)$. By the chain rule, the corresponding weight and bias
gradients are
\begin{equation}
    \frac{\partial \loss_j}{\partial \W^{a}_{i}}
    =
    \frac{\partial \loss_j}{\partial z^{a}_{i,j}}
    \frac{\partial z^{a}_{i,j}}{\partial \W^{a}_{i}}
    =
    \frac{\partial \loss_j}{\partial z^{a}_{i,j}}
    \x_j
    \mathbb{1}_{z^{a}_{i,j}>0},
\end{equation}
and
\begin{equation}
    \frac{\partial \loss_j}{\partial b^{a}_{i}}
    =
    \frac{\partial \loss_j}{\partial z^{a}_{i,j}}
    \frac{\partial z^{a}_{i,j}}{\partial b^{a}_{i}}
    =
    \frac{\partial \loss_j}{\partial z^{a}_{i,j}}
    \mathbb{1}_{z^{a}_{i,j}>0}.
\end{equation}
Thus, whenever neuron $i$ is active and its bias gradient is non-zero,
the input can be recovered exactly as
\begin{equation}
    \x_j
    =
    \frac{\partial \loss_j}{\partial \W^{a}_{i}}
    \left(
        \frac{\partial \loss_j}{\partial b^{a}_{i}}
    \right)^{-1}.
\end{equation}

If the attacked layer is the first layer of the network, this expression
reveals the original training sample, otherwise, it reveals the
intermediate representation supplied as input to that layer. The
reconstruction is obtained in closed form and does not require an
iterative optimization procedure. When gradients are aggregated over a
batch, however, the same ratio generally produces a weighted linear
combination of all samples activating the neuron. Disentangling these
contributions is therefore the central problem addressed by subsequent
analytical attacks.

In the honest-but-curious setting,
\citeA{zhuRGAPRecursiveGradient2021} extended closed-form inversion from
shallow affine models to deep fully connected and convolutional networks
through a recursive, layer-wise reconstruction procedure. R-GAP also
introduced a rank-based analysis to identify the layers and
architectures for which exact recovery is possible, although its
strongest results concern single-example gradients and networks
satisfying suitable rank and invertibility conditions.
\citeA{pan_gia} subsequently characterized batch reconstruction through
the notion of \emph{neuron exclusivity}. A ReLU neuron is exclusive to a
sample $\x_j$ when it is activated by $\x_j$ but remains inactive for
every other sample in the batch. In this case, its aggregated weight and
bias gradients contain only the contribution of $\x_j$, enabling its
closed-form recovery. \citeA{luAPRILFindingAchilles2022} showed that
analytical inversion can also be extended to self-attention layers and
that learnable positional embeddings may expose the ordering of the
reconstructed image patches. In a complementary theoretical setting,
\citeA{wang2023reconstructingtrainingdatamodel} established
identifiability guarantees for sufficiently wide shallow and deep
networks and proposed a tensor-decomposition method for reconstructing training samples from a gradient evaluated at suitably selected model parameters. These early passive attacks were nevertheless limited by restrictive conditions, such as single-samples gradients, or natural occurrence of the exclusive activations. Then, SPEAR~\cite{dimitrovSPEARExactGradient2024} addressed this limitation by
exploiting the low-rank factorization of fully connected-layer gradients
together with the sparsity induced by ReLU activations. This enabled the
exact reconstruction of complete batches of up to approximately 25
high-dimensional images, using dictionary learning techniques~\cite{pmlr-v23-spielman12}. However, its sampling procedure has exponential complexity in the batch size, therefore, \cite{bakarsky2025spearscalinggradientinversion} improved the complexity limitation by replacing the search with more efficient dictionary learning techniques. For language models, DAGER~\cite{petrovDAGERExactGradient2024}
built on the same low-rank intuition as SPEAR: gradients computed over a
batch span a low-dimensional subspace containing the private inputs.
Instead of using ReLU-induced sparsity to disaggregate continuous
samples, DAGER leverages the finite vocabulary and known embedding
matrix to identify tokens lying in the gradient-induced subspace and
recover complete textual batches. \cite{drencheva2025grain} extend the same underlying theory to GNN layers, recovering both client input features and graph structure, under the assumption of discrete input features.

A distinct line of work considers a malicious server that is allowed to
manipulate the model distributed to clients. A seminal example is
\textit{Robbing the Fed}~\cite{fowlRobbingFedDirectly2021}, which
introduced an \emph{imprint module} consisting of a maliciously
configured fully connected ReLU layer. Its ordered activation thresholds
separate the contributions of different samples across consecutive
neurons, allowing the attacker to isolate individual samples from the
corresponding weight and bias gradients. The private inputs can then be
reconstructed in closed form from a single communication round gradients.

In parallel, \citeA{pasquiniEludingSecureAggregation2022a} showed that a
malicious server can circumvent Secure Aggregation
(SA)~\cite{secure_aggregation_bonawitz}. The server sends an honest model to the target client
and manipulated models to the remaining clients, forcing the latter to
produce zero gradients. As a result, the secure aggregate reveals the
target client's individual update. The imprint module subsequently became a fundamental building block for several attacks. Decepticons~\cite{fowlDecepticonsCorruptedTransformers2022} combines an
imprint-style linear leakage module with malicious modifications of the
attention layers, allowing the attacker to associate the recovered
tokens with their positions and reconstruct the original sequences in
transformer-based language models.
Panning for Gold~\cite{chu2023panning} further adapts this construction
from indiscriminate reconstruction to the targeted extraction of
privacy-sensitive information. Modified attention heads tag sequences
containing an attacker-defined trigger, while the linear leakage module
filters out irrelevant tokens and exposes only the targeted
continuations. \citeA{Zhao_cvf} examined the resource overhead of linear-layer leakage,
showing that naively scaling the imprint module with the total number of
aggregated samples leads to prohibitive model growth. They proposed a
sparse construction that separates the scaling with the number of
clients from that with the number of local samples, thereby reducing the
number of injected parameters.  This line of work was subsequently
extended by LOKI~\cite{zhaoLOKILargescaleData2023a}, which combines
client-specific convolutional mappings with fully connected imprint
layers. The convolutional mappings keep the gradient contributions of
different clients separated after aggregation, while the imprint layers
isolate individual samples within each client. Scale-MIA~\cite{shiScaleMIAScalableModel2025} applies the same imprint
mechanism to an intermediate layer of the target model. Rather than
recovering the original inputs directly, the malicious linear layers
isolate and reconstruct their latent representations in closed form,
which are subsequently mapped back to the input space using a
generative decoder trained on auxiliary data.

An alternative ReLU-leakage attack was proposed by~\citeA{boenischWhenCuriousAbandon2023} through the \emph{Trap Weights} attack. initializations may naturally cause some neurons in a fully connected ReLU layer to activate for only one sample, thereby exposing
that sample through linear-layer leakage. Trap Weights amplifies this
phenomenon by inconspicuously rescaling the weights of existing layers,
increasing the number of neurons that isolate individual samples without
modifying the model architecture. A theoretical bound on the
reconstruction capability of this attack is discussed in
Chapter~\ref{chap:ctp}. A related approach was proposed
by~\citeA{ZHANG2023119421}, who train malicious parameters for an
existing fully connected layer using an auxiliary dataset. Their method
extends analytical extraction beyond ReLU networks to models employing
Tanh and Sigmoid activations. A different architecture-preserving strategy is adopted by
MKOR~\cite{wangMaximumKnowledgeOrthogonality2024}. The attack
maliciously modifies the parameters of existing convolutional and fully
connected layers to maximize the orthogonality between prior knowledge
about the input structure and the information revealed by the gradients.
For images, this prior is based on the assumption that neighbouring
pixels tend to have similar intensities. Under this assumption, MKOR
enables analytical reconstruction of multiple samples from large
batches without requiring auxiliary data or changes to the model
architecture.

Building on this line of work, Chapter~\ref{chap:ctp} introduces a
geometric perspective on analytical gradient leakage and proposes an
attack capable of recovering all samples in a batch. Chapter~\ref{chap:vgia} then
investigates the previously overlooked problem of verifiability and
introduces a method that certifies the exact isolation of a private
sample before reconstructing it analytically.



\subsection{Other Attacks}
Federated learning is subject to several additional forms of privacy leakage that are partially related to those discussed in the previous subsection. Source inference attacks, for instance, are closely related to membership inference attacks (MIAs). Rather than merely determining whether a sample belongs to a client’s training dataset, they seek to identify the client from which a known or suspected training record originated~\cite{hu2021sourceinferenceattacksfederated, Hu2024sia, ijcai2025p536}. Property inference attacks are instead more closely related to attribute inference attacks (AIAs). However, rather than recovering a specific feature of an individual record, they aim to infer statistical or semantic characteristics of a client’s local dataset, such as the presence or prevalence of a sensitive group or concept, including properties unrelated to the intended learning task. Examples include the presence of people wearing glasses in an image dataset or of particular medical conditions in a patient dataset~\cite{melis2018exploitingunintendedfeatureleakage, Zhang_pia, Wang_pia, Kerkouche_pia}. Another category of attacks targets graph topology, either by exploiting the communication graph that connects participants in decentralized learning or by recovering the topology of private graph-structured data. In the former setting, adversaries may exploit their position in a peer-to-peer network and the gossip-averaging process to reconstruct information belonging to neighbouring or even non-neighbouring users~\cite{Pasquini_insecurity, ElMrini_graph}. Finally, canary-based attacks and privacy audits deliberately introduce crafted examples, gradients, or client updates and subsequently test whether their participation can be detected. These methods provide empirical estimates of membership leakage or differential-privacy loss, rather than inferring an unknown, naturally occurring property of another client’s data~\cite{maddock2023canife, Nasr_tight_auditing, andrew2024oneshot}. Collectively, these attack families demonstrate that withholding raw data is not, by itself, sufficient to guarantee privacy. The next subsection therefore provides an overview of the main defense mechanisms proposed in the literature.
\subsection{Privacy Mechanisms in Federated Learning}
Several mechanisms have been proposed to mitigate privacy risks in federated learning. As discussed by~\citeA{kairouz_open_problems}, privacy should not be regarded as a binary property provided by a single mechanism. Instead, effective privacy-preserving systems typically adopt a layered approach, combining complementary mechanisms that address different leakage channels, adversarial capabilities, and trust assumptions. This subsection provides an overview of the main privacy-preserving mechanisms used in federated learning. For more comprehensive discussions, the reader is referred to dedicated surveys on privacy and security in federated learning~\cite{Rao_privacy_survey, Yin_defense_taxonomy, sok_secure_aggregation}.

Differential privacy (DP)~\cite{dp_dwork,dwork_dp_book} is widely regarded as the de facto standard for providing formal and quantifiable privacy guarantees. Rather than directly measuring the amount of information disclosed by an algorithm, DP bounds the extent to which the distribution of its outputs can change when the contribution of a single sample is added to or removed from the input dataset. More formally, a mechanism $\mathcal{M}$ satisfies $(\varepsilon, \delta)$-differential privacy guarantees if, for every pair of adjacent datasets $\mathcal{D}$ and $\mathcal{D}'$, and for every measurable set of outputs $S \subseteq \mathrm{Range}(\mathcal{M})$,
\begin{equation}\label{eq:dp}
    \mathrm{Pr}\left[ \mathcal{M}(\mathcal{D} \in S)\right] \leq e^{\varepsilon} \mathrm{Pr}\left[ \mathcal{M}(\mathcal{D}' \in S)\right] + \delta.
\end{equation}
Here $\varepsilon$ controls the multiplicative difference between the two output distributions, and $\delta$ is a small probability of failure. In practice, $\varepsilon$ quantifies the maximum amout of information that the output of the algorithm can disclose, therefore a smaller value indicates a stronger privacy guarantee. A common mechanism for training neural networks with record-level DP is DP-SGD~\cite{Abadi_dp}. At every optimization step, DP-SGD clips each per-example gradient to bound its sensitivity and adds calibrated Gaussian or Laplacian noise to the aggregated gradient. 

In Federated Learning, depending on where the perturbation is applied and which parties are trusted, different DP deployment models can be distinguished. Under \emph{local differential privacy} (LDP), each client randomizes its contribution before transmitting it to the server. Consequently, the server never observes an unperturbed update, and the privacy guarantee does not rely on trusting the server with individual client contributions. However, because each transmitted contribution must satisfy DP independently, LDP generally requires substantially more noise than mechanisms that operate on aggregated updates. This can adversely affect model convergence and final-model utility, particularly for high-dimensional updates, small client populations, or strong privacy requirements~\cite{geyer2018differentiallyprivatefederatedlearning,kairouz_open_problems}.

Under \emph{central differential privacy} (CDP), a trusted server or aggregator receives the individual client updates, bounds their sensitivity through clipping, computes their aggregate, and adds calibrated noise before using or releasing it. Since noise is added only once to the aggregate, CDP typically provides a more favourable privacy--utility trade-off than LDP. Nevertheless, conventional CDP requires the aggregator to be trusted because it observes individual updates before the privacy mechanism is applied. It therefore does not protect clients against a malicious or compromised server~\cite{mcmahan_user_dp,kairouz_open_problems}.

To improve utility relative to LDP without exposing individual unperturbed updates, \emph{distributed differential privacy} (DDP) combines client-side randomization with secure aggregation~\cite{kairouz_discrete_gaussian, agarwal2021the, chen2022poisson}. Secure aggregation is a cryptographic multiparty-computation primitive that allows the server to recover only an aggregate, typically the sum of the clients' updates, while hiding each individual contribution~\cite{secure_aggregation_bonawitz,sok_secure_aggregation}. In DDP, each client clips its update and contributes a share of the noise required to achieve a target DP guarantee. Although an individual contribution does not necessarily satisfy LDP, the securely aggregated sum can provide a central-DP-like guarantee because the server observes only the noisy aggregate and cannot separate the contributions of individual clients~\cite{kairouz_discrete_gaussian, agarwal2021the, chen2022poisson}.

Distributed DP introduces additional assumptions and system costs. Its guarantees depend on sufficient client participation and on bounds regarding dropout and collusion, while secure aggregation adds computational, communication, and coordination overheads~\cite{kairouz_discrete_gaussian, chen2022fundamentalpricesecureaggregation}. Moreover, secure aggregation hides individual updates but does not verify the model distributed by the server or the validity of client contributions. In some settings, a malicious server may therefore circumvent its protection through inconsistent or manipulated models, without breaking the underlying cryptographic protocol~\cite{pasquiniEludingSecureAggregation2022a,zhaoLOKILargescaleData2023a}.

More generally, DP entails an inherent privacy-utility trade-off: stronger guarantees may reduce convergence and accuracy, while large values of $\varepsilon$ can provide limited practical protection~\cite{Jayaraman_privacy_evaluation,cummings2024advancingdifferentialprivacyfuture}. To some extents, empirical attacks may still succeed when privacy parameters are weak, as we will examin in Chapter~\ref{chap:aia}.


Beyond formal privacy guarantees, several attack-specific empirical countermeasures have been proposed in the literature~\cite{bai_survey_mai,carletti2025sok}. For membership and property inference, researchers have investigated selective gradient sharing, model perturbation, and weight pruning~\cite{melis2018exploitingunintendedfeatureleakage,yang_perturbation,Stripelis_pruning}. However, the protection offered by these techniques is generally specific to the attacks and threat models considered during their evaluation.

Gradient inversion attacks have motivated defenses operating at different stages of the learning process. Data-level approaches transform or replace sensitive samples before computing the shared updates~\cite{jeterOASISOffsettingActive2024,fanRefinerDataRefining}. Other methods obfuscate intermediate representations or model updates through updates perturbation~\cite{sunSoteriaProvableDefense2021, Xue_pruning, zhangCENSORDefenseGradient2025}, model modifications~\cite{scheligaPRECODEGenericModel2022}, learning algorithm modifications~\cite{Wudefense2024, Ye_privacy2023}. Client-side detection mechanisms have also been proposed to identify model manipulations performed by malicious servers before local training begins~\cite{wangHearNoEvil2025,Carletti_2026}.

Nevertheless, these methods generally provide empirical, attack-dependent protection rather than universal privacy guarantees. Their effectiveness depends on the adversary's knowledge and capabilities, as well as on the model architecture, training configuration, and evaluation setting. In particular, adaptive attacks have been shown to circumvent gradient compression, pruning, stochastic defenses, and noise-based perturbation~\cite{Yue_gia2023,scheligaCombiningStochasticDefenses2025,liu2025mjolnirbreakingshieldperturbationprotected,mengEnhancedPrivacyLeakage2025,carletti2025sok}. These limitations motivate further investigation of the information exposed by federated learning protocols and of the conditions under which such information can be exploited. 

\section{Summary of the Thesis Contribution}

The preceding sections have shown that federated learning changes, rather than eliminates, the trust assumptions underlying machine learning. The central server retains a privileged position in the training process, as it coordinates client participation, distributes model parameters, and observes the resulting updates. It is therefore essential to understand how this position may be exploited, since privacy should not depend solely on the assumption that the server will act in the interests of participating clients.

Strong privacy attacks play an important role in this assessment. They reveal information that remains exposed by the training protocol, establish empirical lower bounds on the leakage achievable by an adversary, and provide demanding baselines for evaluating privacy-preserving mechanisms.

Motivated by these considerations, this thesis pursues two main objectives:
\begin{enumerate}
    \item To deepen the understanding of privacy leakage in federated learning
    from the perspective of the central server, characterizing what information
    can be inferred from clients' updates under both honest-but-curious and
    malicious-server threat models.

    \item To develop stronger, more scalable, and more reliable privacy attacks
    that improve the empirical auditing of federated learning systems and
    provide more informative baselines for the design and evaluation of
    privacy-preserving mechanisms.
\end{enumerate}

These contributions are motivated by a constructive objective. Federated learning can meaningfully contribute to preserving privacy, but the protection it provides should not depend on the server’s goodwill alone. Its effectiveness ultimately depends on how the training protocol is designed, implemented, and complemented by appropriate safeguards. By clarifying what information a server may infer under different threat models, this thesis aims to help researchers, system designers, organizations, and service providers make informed design choices and assess privacy claims more rigorously. Based on these motivations, the remainder of this section summarizes the contributions of the thesis.

\subsection*{Contribution \#1: Attribute Inference Attacks for Federated Regression Tasks.}

\noindent\textbf{Problem:}
Attribute inference attacks; federated regression; empirical privacy auditing.

\medskip
\noindent\textbf{Research gap:}
Existing attribute inference attacks for federated learning have predominantly
focused on classification tasks and rely on matching the gradients generated by
candidate sensitive attributes with the updates observed during training.
Their effectiveness in regression settings had remained largely unexplored.
Moreover, model-based attribute inference attacks, originally developed for
centralized learning, had not been adapted to exploit the information exchanged
during federated training. As a result, the poor performance of existing
gradient-based attacks could incorrectly suggest that federated regression
provides stronger privacy protection.

\noindent\textbf{Main contributions:}
\begin{itemize}
    \item We establish a theoretical lower bound on the accuracy of model-based
    attribute inference attacks for least-squares regression, showing how the
    attack success depends on the local model's training error and on the
    influence of the sensitive attribute on the prediction.

    \item We introduce model-based attribute inference attacks for federated
    regression under both honest-but-curious and malicious-server threat
    models. The proposed methods first approximate the optimal local model of
    the targeted client and then use it to infer the sensitive attributes of
    individual records.

    \item Through experiments on real-world datasets, we show that the proposed
    attacks substantially outperform existing gradient-based approaches,
    particularly under heterogeneous client distributions, and provide stronger
    empirical baselines for evaluating privacy leakage in federated regression.
\end{itemize}

\medskip
\noindent\textbf{Publication:}
 \textbf{Francesco Diana}, Othmane Marfoq, Chuan Xu, Giovanni Neglia,
    Frédéric Giroire, and Eoin Thomas.
    ``Attribute Inference Attacks for Federated Regression Tasks.''
    In \emph{Proceedings of the AAAI Conference on Artificial Intelligence},
    vol.~39, no.~15, pp.~16271--16279, 2025.



\subsection*{Contribution \#2: A Hyperplane-Based
Data Reconstruction Attack in Federated Learning.}

\noindent\textbf{Problem:} Gradient inversion attacks; malicious federated-learning servers; reconstruction
of large and low-dimensional data batches.

\medskip
\noindent\textbf{Research gap:}
Existing gradient inversion attacks either rely on auxiliary knowledge about the
clients' data distribution or lose effectiveness as the number of records
contributing to an update increases. In particular, sparsity-based analytical
attacks attempt to configure neurons so that individual records can be separated
from the remaining batch. Although effective for some high-dimensional image
data, their performance degrades considerably for large batches and
low-dimensional tabular records. Consequently, prior attacks could not reliably
recover all the records contained in large aggregated updates.

\noindent\textbf{Main contributions:}
\begin{itemize}
    \item We provide a geometric interpretation of existing sparsity-based
    analytical attacks and derive theoretical upper bounds on their
    reconstruction capabilities, explaining their limited effectiveness for
    large batches and low-dimensional data.

    \item We introduce a hyperplane-based reconstruction attack in which a
    malicious server controls the contribution of client records to the
    observed gradients and progressively isolates them across communication
    rounds, without requiring auxiliary data or prior knowledge of the client
    data distribution.

    \item We show theoretically and empirically that the proposed attack can
    recover complete batches independently of their size, given sufficiently
    many communication rounds, and demonstrate perfect reconstruction of image
    and tabular batches containing up to \(4{,}096\) records.
\end{itemize}

\medskip
\noindent\textbf{Publication:}
 \textbf{Francesco Diana}, André Nusser, Chuan Xu, and Giovanni Neglia.
    ``Cutting Through Privacy: A Hyperplane-Based Data Reconstruction Attack
    in Federated Learning.''
    In \emph{The 41st Conference on Uncertainty in Artificial Intelligence},
    2025.


\subsection*{Contribution \#3: No More Guessing: A Verifiable Gradient
Inversion Attack in Federated Learning}

\noindent\textbf{Problem:}
Verifiability of gradient inversion attacks; certified record reconstruction;
privacy auditing without access to ground-truth data.

\medskip
\noindent\textbf{Research gap:}
Gradient inversion attacks are generally evaluated by comparing reconstructed
records with the original client data. Such an evaluation is available to the
researcher conducting an experiment, but not to an actual adversary, who does
not know the private batch. Existing attacks may therefore not distinguish
correct records from incorrect ones. Visual inspection can sometimes provide a
heuristic assessment for images or text, but it is costly for large batches and
is largely ineffective for numerical tabular data. Moreover, the geometric
search introduced in the previous contribution relies on a manually selected
stopping threshold and cannot determine whether a non-empty region contains one
record or several records.

\medskip
\noindent\textbf{Main contributions:}
\begin{itemize}
    \item We introduce a verifiable gradient inversion attack that uses an
    algebraic subspace test to certify when records have been individually
    isolated, providing an intrinsic guarantee of reconstruction correctness
    without requiring access to the original client data.

    \item We develop an adaptive hyperplane-search procedure that analytically
    reconstructs certified inputs, recovers their corresponding targets, and
    reallocates subsequent queries to unresolved regions, thereby avoiding
    heuristic stopping thresholds and unnecessary communication rounds.

    \item We validate the attack on tabular and image datasets for both
    regression and classification tasks, demonstrating exact and automatically
    certified recovery under large-batch aggregation in settings where previous
    attacks either fail or cannot assess reconstruction correctness.
\end{itemize}

\medskip
\noindent\textbf{Publication:}
    \textbf{Francesco Diana}, Chuan Xu, André Nusser, and Giovanni Neglia.
    ``No More Guessing: A Verifiable Gradient Inversion Attack in Federated
    Learning.''
    \emph{arXiv preprint arXiv:2604.15063}, 2026.

\section{Publications}
\subsection*{Published}
\noindent\begin{itemize}
    \item  \textbf{Francesco Diana}, Othmane Marfoq, Chuan Xu, Giovanni Neglia,
Frédéric Giroire, and Eoin Thomas.
``Attribute Inference Attacks for Federated Regression Tasks.''
In \emph{Proceedings of the AAAI Conference on Artificial Intelligence},
vol.~39, no.~15, pp.~16271--16279, 2025.
    \item \textbf{Francesco Diana}, André Nusser, Chuan Xu, and Giovanni Neglia.
``Cutting Through Privacy: A Hyperplane-Based Data Reconstruction Attack
in Federated Learning.''
In \emph{The 41st Conference on Uncertainty in Artificial Intelligence},
2025.
\end{itemize}

\subsection*{Under Review}
\noindent\begin{itemize}
    \item \textbf{Francesco Diana}, Chuan Xu, André Nusser, and Giovanni Neglia.
    ``No More Guessing: A Verifiable Gradient Inversion Attack in Federated
    Learning.''
    \emph{arXiv preprint arXiv:2604.15063}, 2026.
\end{itemize}

\section{Thesis Outline}
The remainder of this thesis is organized as follows. Chapter~\ref{chap:aia}
studies attribute inference attacks for federated regression tasks. It introduces
model-based attacks for both honest-but-curious and malicious adversaries,
provides theoretical guarantees for least-squares regression, and evaluates
their effectiveness against existing gradient-based approaches.
Chapter~\ref{chap:ctp} turns to the complete reconstruction of clients' private
records. It provides a geometric analysis of the limitations of existing
analytical attacks and introduces a hyperplane-based method capable of
reconstructing large batches of image and tabular data.
Chapter~\ref{chap:vgia} addresses the problem of reconstruction verifiability
and proposes an attack that certifies when a private record has been correctly
isolated and recovered, while adaptively reducing the number of communication
rounds required by the reconstruction process. Finally, Chapter~\ref{chap:conclusion}
summarizes the main findings of the thesis and discusses possible directions and challenges
for future research on privacy attacks in federated
learning.
